---
title: "Formative Assessment 4"
output: pdf_document
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-2cm}   % move title up
geometry: top=2cm, bottom=2cm, left=2cm, right=2cm
mainfont: "Georgia"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(readxl)
  library(tidyverse)
  library(ggplot2)
  library(car)
  library(PerformanceAnalytics)
  library(xts)
  library(rugarch)
  library(quantmod)
  library(tseries)
  library(forecast)
})
```

# Stock Returns

```{r}
stock <- read_csv('dataset_stock_returns.csv', show_col_types = FALSE)
stock$dates <- seq.Date(from = as.Date("2010-01-01"), by="days", length.out = nrow(stock))
head(stock)

```

## Visual Exploration of Volatility

### Return Series and Rolling Standard Deviation
```{r,fig.width=7, fig.height=4}
plot(stock$dates, stock$returns, type = "l", col = "blue",
     main = "Returns Series", xlab = "Time", ylab = "Return")
```

The return series plot shows returns fluctuating randomly around 0, with no apparent trend or seasonality and showing alternating periods of high and low variability. Notably, two distinct time intervals stand out where returns exhibit large negative and positive movements, preceded and followed by periods of relatively small changes. This pattern is indicative of volatility clustering.

```{r, fig.width=6, fig.height=4}
hist(stock$returns, main="Distribution of Stock Returns", 
     col = "skyblue", xlab= "Return", freq=FALSE)

x <- seq(min(stock$returns, na.rm = TRUE),
         max(stock$returns, na.rm = TRUE),
         length = 100)

lines(x,dnorm(x,mean = mean(stock$returns, na.rm = TRUE),
              sd = sd(stock$returns, na.rm = TRUE)), col="red",lwd=2)
```

The plot of returns is not symmetrically distributed, rather it is negatively skewed. This indicates that large negative returns occur more frequently than large positive ones. Compared to the normal distribution, the returns have moderate kurtosis, meaning that extreme changes happen moderately.

```{r, fig.width=6, fig.height=7}

par(mfrow=c(2,1))
rolling_sd <- runSD(stock$returns, n = 20, cumulative = FALSE)
plot(stock$dates, rolling_sd, type = "l", col = "red", 
     main = "Rolling Standard Deviation (n=20)", xlab = "Time", ylab = "Volatility")
grid()
plot(stock$dates, stock$returns, type = "l", col = rgb(red = 0, green = 0, blue = 1, alpha = 0.5), 
     main = "Returns and Rolling Standard Deviation (n=20)",
     xlab = "Time", ylab = "Value")
lines(stock$dates, rolling_sd, col = "red", lwd = 2)
grid()

```

Assuming 20 trading days per month, the rolling standard deviation line shows the monthly volatility of the stock returns. It rises sharply during turbulent periods, reflecting increased volatility, and falls during calm periods when the market stabilizes. This behavior suggests that the variance of returns is not constant over time, implying conditional heteroskedasticity.


**Plot ACF of Squared Returns**
```{r, fig.width=6, fig.height=4}
acf((stock$returns)^2, main = "ACF of Squared Returns")
```

The ACF of squared returns drops quickly to zero and is insignificant at lags 1–2 but becomes significant again at lag 3. This pattern suggests an absence of immediate volatility clustering, but the presence of some delayed or periodic dependence in volatility roughly every three periods. The changing variance of returns over time indicates conditional heteroskedasticity, consistent with the presence of volatility dynamics. This could point to a potential ARCH(3)-type effect, where volatility depends on shocks occurring three periods earlier.

```{r}
library(FinTS)
ArchTest(stock$returns, lags = 12)
```

## Model Estimation

### Model 1: ARCH(3)

The ACF of squared returns shows a small but noticeable dependence up to lag 3, suggesting that volatility today may be influenced by shocks over the past three periods. The ARCH(3) model directly captures this short-term volatility dependence by including three lag terms.

```{r, fig.width=6, fig.height=4}
arch3_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(3,0)),
  mean.model     = list(armaOrder = c(0,0), include.mean = TRUE),
  distribution.model = "std")

arch3_fit <- ugarchfit(spec = arch3_spec, data = stock)
```

Coefficients and Standard Errors
```{r}
arch3_fit@fit$matcoef
```

The mean $\mu$ term is insignificant, suggesting that daily stock returns fluctuate randomly around zero, reflecting the uncertainty in market. The constant term $\omega$ is small but highly significant (p < 0.001), indicating a low baseline level of market volatility during stable periods. Among the three ARCH coefficients, only the third lag ($\alpha_3=0.107, p = 0.009$) is statistically significant.This suggests that volatility is primarily influenced by shocks from three days prior, indicating a delayed reaction to past shocks rather than an immediate response. The shape parameter is also insignificant, suggesting that the residuals do not strongly deviate from normality. These results imply that market participants remain cautious or risk-averse for some days following major movements. 

### Model 2: GARCH(1,1)

This model model extends the original ARCH framework by allowing current volatility to depend on recent shocks (ARCH term) and recent volatility (GARCH term). 

```{r}
garch11_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
  distribution.model = "std"
)

garch11_fit <- ugarchfit(spec = garch11_spec, data = stock)
```

Coefficient Estimates and Standard Errors
```{r}
garch11_fit@fit$matcoef
```

Again, the $\mu$ term is insignificant in this model, while the constant $\omega$ term is small and is only marginally significant (p < 0.1), implying a low but nonzero long-run variance that forms the baseline level of market volatility. The ARCH coefficient ($\alpha_1=0.054, p=0.01$) is positive and significant, suggesting that recent shocks have a considerable effect on current volatility. Comparably, the GARCH coefficient ($\beta_1=0.837, p < 0.001$) is relatively large and significant, indicating that past volatility strongly persists into the current period. Together, the ARCH and GARCH terms imply that volatility is highly persistent with shocks dissipating after some days. This reflects investors’ cautious behavior, delayed market reactions, and ongoing uncertainty after large movements.
 


### AIC Comparison
```{r}
data.frame(
  Model = c("ARCH(3)", "GARCH(1,1)"),
  AIC = c(infocriteria(arch3_fit)[1], infocriteria(garch11_fit)[1]),
  BIC = c(infocriteria(arch3_fit)[2], infocriteria(garch11_fit)[2])
)
```

While both models have very similar AIC and BIC values, GARCH(1,1) has slightly lower values, indicating a marginally better fit. Furthermore, GARCH(1,1) is able to capture both short-term shock effects and persistent volatility, aligning with the observed volatility clustering in the returns series and volatility plots. Meanwhile,ARCH(3) only captures shocks up to three days ago and misses the longer-term persistence, which is why its fit is slightly worse.

## Diagnostics and Volatility Interpretation

### COnditional Volatility

```{r, fig.width=7, fig.height=4}
plot(sigma(arch3_fit), main= "Conditional Volatility from ARCH(3)",
        ylab="Volatility", col="blue", lwd=1)
grid()
```


```{r, fig.width=7, fig.height=4}
plot(sigma(garch11_fit), main= "Conditional Volatility from GARCH(1,1)",
        ylab="Volatility", col="blue", lwd=1)
grid()
```

As ARCH(3) only relies on past shocks, it produced a more jagged volatility series. It responds more sharply to individual shocks and drops back to the baseline quickly, which may underestimate volatility. In contrast, the GARCH(1,1) provides a smoother visualization of volatility decaying realistically, demonstrating the persistence of volatility. 

### ACF of Squared Residuals

```{r, fig.width=7, fig.height=4}
arch3_resid <-residuals(arch3_fit, standardize=TRUE)
garch11_resid <-residuals(garch11_fit, standardize=TRUE)
par(mfrow=c(1,2), mai = c(1, 0.8, 0.7, 0.5))
acf(arch3_resid^2, main ="ACF of ARCH(3) Residuals^2")
acf(garch11_resid^2, main ="ACF of GARCH(1,1) Residuals^2")
```

ARCH-LM test
```{r}
ArchTest(arch3_resid)
ArchTest(garch11_resid)
```

The ACF plots show no remaining ARCH effect in the residuals, indicating a good fit for both models. The ARCH-LM test was also performed to formally assess the presence of ARCH effects in the residuals. Results for both models yielded p-values above the standard significance threshold of 0.05, confirming the findings from the ACF plots.


Because stock returns typically exhibit long-lasting volatility clustering, the GARCH(1,1) model offers a more realistic representation of the underlying volatility dynamics over the simple ARCH model.


## Volatility Forecasting

### Forecast conditional volatility for 10 future periods 
```{r}
arch3_fc <- ugarchforecast(fitORspec = arch3_fit, n.ahead=10)
garch11_fc <- ugarchforecast(fitORspec =garch11_fit, n.ahead=10)

arch3_fc_vol <- sigma(arch3_fc)
garch11_fc_vol <- sigma(garch11_fc)

fc <- data.frame(cbind(arch3_fc_vol,garch11_fc_vol))
colnames(fc) <- c("ARCH(3)", "GARCH(1,1")
fc
```


### Forecasted volatility path with confidence intervals

ARCH(3)
```{r}
# arch
fc <- arch3_fc
hist_vol <- tail(sigma(arch3_fit),100)
fc_vol <- sigma(fc)
fc_mean <- fitted(fc)
hist_dates <- index(hist_vol)
fc_dates <- seq(from=hist_dates[length(hist_dates)]+1, by ="days",
                length.out =10)

upper <- fc_vol*1.96
lower <- fc_vol*-1.96

all_vol <- c(as.numeric(hist_vol), fc_vol)
all_dates <- c(hist_dates, fc_dates)

plot(all_dates, all_vol, type = "n", col = "blue", lwd = 2,
     ylim = range(c(lower, upper, fc_vol)),
     main = "10-Day Forecasted Conditional Volatility",
     xlab = "Time", ylab = "Volatility")


polygon(c(fc_dates, rev(fc_dates)),
        c(upper, rev(lower)),
        col = "lightblue", border = NA)

lines(hist_dates, hist_vol, col="black")

lines(fc_dates, fc_vol, col="blue", lty=2)

legend("bottomleft", legend = c("Historical Data", "Forecast"), 
       col = c("black", "blue"), lwd = c(1,1), lty = c(1,2))

```

what happens to volatility over the forecast horizon (does it decay, persist, or show asymmetry?)

```{r}
library(xts)
hist <- sigma(arch3_fit)

n_hist <- length(hist)
n_fc   <- length(arch3_sigma)

# Generate date sequences
hist_dates <- seq(as.Date("2020-01-01"), by = "days", length.out = n_hist)
fc_dates   <- seq(hist_dates[n_hist] + 1, by = "days", length.out = n_fc)

# Combine volatility vectors and dates
all_vol <- c(hist, arch3_sigma)
all_dates <- c(hist_dates, fc_dates)

# Create xts object
vol_xts <- xts(all_vol, order.by = all_dates)

```



Reflection and Application

Discuss in which contexts (finance, energy, currency) volatility modeling is most relevant.

Identify any limitations or possible model improvements.



# Bitcoin Returns

```{r}
bitcoin <- read_csv('dataset_bitcoin_returns.csv', show_col_types = FALSE)
head(bitcoin)
```

## Visual Exploration of Volatility

### Return Series and Rolling Standard Deviation
```{r,fig.width=7, fig.height=4}
plot(1:nrow(bitcoin), bitcoin$returns, type = "l", col = "blue",
     main = "Bitcoin Returns Series", xlab = "Time", ylab = "Return")
```

```{r, fig.width=6, fig.height=4}
hist(bitcoin$returns, main="Distribution of Bitcoin Returns", 
     col = "skyblue", xlab= "Return", freq = FALSE)

x <- seq(min(bitcoin$returns, na.rm = TRUE),
         max(bitcoin$returns, na.rm = TRUE),
         length = 100)

lines(x,dnorm(x,mean = mean(bitcoin$returns, na.rm = TRUE),
              sd = sd(bitcoin$returns, na.rm = TRUE)), col="red")
```


```{r, fig.width=7, fig.height=7}
par(mfrow=c(2,1))
plot(1:nrow(bitcoin), rolling_sd, type = "l", col = "red", 
     main = "Rolling Standard Deviation (n=20)", xlab = "Time", ylab = "Volatility")
rolling_sd <- runSD(bitcoin$returns, n = 20, cumulative = FALSE)
plot(1:nrow(bitcoin), bitcoin$returns, type = "l", col = "blue",
     main = "Returns and Rolling Standard Deviation (n=20)",
     xlab = "Time", ylab = "Value")
lines(1:nrow(bitcoin), rolling_sd, col = "red", lwd = 2)
```

```{r, fig.width=6, fig.height=4}
acf(bitcoin^2, main = "ACF of Squared Returns")
```

Slow decay, pattern




## Model Estimation

### Model 1 GARCH(1,1)
```{r}
garch11_spec <-  ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
  distribution.model = "norm"
)

garch11_fit <- ugarchfit(garch11_spec, data = bitcoin$returns)
```

```{r}
garch11_fit@fit$matcoef
```
```{r}
infocriteria(garch11_fit)
```


### Model 2

Record estimated parameters, standard errors, and information criteria (AIC, BIC).

## Diagnostics and Volatility Interpretation
Plot conditional volatility and compare across models.

Analyze residuals for remaining ARCH effects (using ACF of residuals squared or ARCH-LM test).

Comment on which model better captures volatility persistence and asymmetry.

## Volatility Forecasting

Forecast conditional volatility for 10 future periods 

forecasted volatility path with confidence intervals

what happens to volatility over the forecast horizon (does it decay, persist, or show asymmetry?)


Reflection and Application

Discuss in which contexts (finance, energy, currency) volatility modeling is most relevant.

Identify any limitations or possible model improvements.